Patient-Specific Neural Field Modeling from Micro-Electrode Data

### 2010-07-30 ltx/data_paper/DataPaper.tex
corrected the time index
correction to beta in Q function
minor changes and commiting to merge
added derivation of Xi variables to suplimentary information

\section{Introduction}
\begin{itemize}
	\item Understanding of macroscopic neurodynamics
	\item Neural field model
	\item Improvements in electrode technology with great spatiotemporal resolution
\end{itemize}
The aim of this paper is to great a patient-specific neural field model, by estimating parameters of a general mathematical model using data acquired from an epilepsy surgical candidate in a clinical setting. 
\section{Method}

@@

\subsection{E-Step}
The unscented Rauch-Tung-Striebel smoother (URTSS)~\cite{Sarkka2010} was used for the state estimation. The URTSS incorporates an unscented Kalman filter (UKF)~\cite{Julier1997, Merwe2003} in a forward step to estimate posterior states, $\hat{\mathbf x}_t^{f}$, followed by a backward step to compute the smoothed state estimates, $\hat{\mathbf x}_t^{b}$. The first and second order moments of the predicted state are captured by propagating the so-called sigma points through the state equation. An additional element to the URTSS algorithm was created to approximate $q(\mathbf{x}_t)$, which is required for the M-step.

@@

Now taking the expectation of the log likelihood function for all time instants gives the $\mathcal Q$-function, which is the lower-bound that needs to be maximised to gain optimal parameter estimates. The $\mathcal Q$-function is written as  

@@

 \mathcal Q(\boldsymbol \theta,\xi,\boldsymbol\theta',\xi')&=\beta+2\Xi_0\boldsymbol\theta+

@@

	\beta &= \sum_{t=0}^{T-1}\mathbf E_{(\theta',\xi')}\left[ \ln \alpha-\mathbf x_{t+1}^\top\boldsymbol\Sigma_w^{-1}\mathbf x_{t+1}\right] \label{eq:def of beta} \\
\Xi_0&=\sum_{t=0}^{T-1}\left[ \mathbf{\hat x}_{t+1}^\top\boldsymbol\Sigma_e^{-1}q(\mathbf{\hat x}_t)+\int_\Omega\boldsymbol \phi^\top(\mathbf r') \mathbf M_t \boldsymbol\Sigma_e^{-1}  \boldsymbol{\Psi}(\mathbf{r}') f'(\boldsymbol \phi^\top(\mathbf r')\mathbf {\hat x}_t) d\mathbf{r}'\right] \\	
\Xi_1&=\mathbf E_{(\theta',\xi')}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t+1}^\top\right]=\sum_{t=0}^{T-1}\left(\mathbf M_t+\mathbf{\hat x}_t\mathbf{\hat x}_{t+1}^\top\right) \\

@@

When taking the derivative of equation~\ref{eq:Voldermont}, the first term described in equation~\ref{eq:def of beta} will will disappear. Also, when we differentiate with respect to $\boldsymbol\theta$ and $\xi$ and equating it to zero we get
 \boldsymbol \theta=\Xi_2^{-\top}\left(\Xi_0^\top-\xi\Xi_3^\top \right)
 \xi=\frac{tr\left\lbrace \Xi_1 \boldsymbol\Sigma_e^{-1}\right\rbrace-\Xi_3\Xi_2^{-\top}\Xi_0^\top}{tr\left\lbrace \Xi_4 \boldsymbol\Sigma_e^{-1}\right\rbrace-\Xi_3\Xi_2^{-\top}\Xi_3^\top}.
\subsection{Discussion}
The second derivative of the $\mathcal Q$-function with respect to $\xi$ is given by
 \frac{\delta^2 \mathcal Q}{\delta \xi^2}=-2 tr \left\lbrace\Xi_1\boldsymbol\Sigma_e^{-1} \right\rbrace
$\Xi_1 $ is positive definite if the state vector is persistently excited and the covariance matrix $\boldsymbol\Sigma_e^{-1}$ is also positive definite therefore the trace of $\Xi_1\boldsymbol\Sigma_w^{-1}$ is guaranteed to be positive so the second derivative is negative and represents a maximum of $\mathcal Q$-function.\\
The second derivative of the $\mathcal Q$-function with respect to $\boldsymbol \theta$ is given by
 \frac{\delta^2 \mathcal Q}{\delta \boldsymbol\theta^2}=-2 \Xi_2 
$\Xi_2$ is positive definite and invertible and the second derivative is negative definite and represents a maximum of $\mathcal Q$- function.

>>>>

\section{Introduction}
\begin{itemize}
	\item Understanding of macroscopic neurodynamics
	\item Neural field model
	\item Improvements in electrode technology with great spatiotemporal resolution
\end{itemize}
The aim of this paper is to great a patient-specific neural field model, by estimating parameters of a general mathematical model using data acquired from an epilepsy surgical candidate in a clinical setting.

Instead of extracting features we aim to extract meaningful parameters...

\section{Method}

@@

\subsection{E-Step}
The unscented Rauch-Tung-Striebel smoother (URTSS)~\cite{Sarkka2010} was used for the state estimation. The URTSS incorporates an unscented Kalman filter (UKF)~\cite{Julier1997, Merwe2003} in a forward step to estimate posterior states, $\hat{\mathbf x}_t^{f}$, followed by a backward step to compute the smoothed state estimates, $\hat{\mathbf x}_t^{b}$. The first and second order moments of the predicted state are captured by propagating the so-called sigma points through the state equation. An additional step is required in the $\mathbf E$-step to calculate the cross-covariance matrix $\mathbf M_{t+1}=\textrm{cov}(\mathbf x_t,\mathbf x_{t+1})$. An additional element to the URTSS algorithm was also created to approximate $q(\mathbf{x}_t)$, which is required for the M-step.

@@

Substituting \ref{eq:qTaylor} in \ref{eq:Qfunctionintrace} and taking the expectation of the log likelihood function for all time instants gives an approximate to the lower-bound that needs to be maximised to gain optimal parameter estimates. Therefore $\mathcal Q$-function can be approximated as  

@@

 \mathcal Q(\boldsymbol \theta,\xi,\boldsymbol\theta',\xi')&\approx\beta+2\Xi_0\boldsymbol\theta+

@@

	\beta &= \sum_{t=0}^{T-1}\mathbf E_{(\theta',\xi')}\left[ \ln \alpha-\mathbf x_{t+1}^\top\boldsymbol\Sigma_e^{-1}\mathbf x_{t+1}\right] \label{eq:def of beta} \\
\Xi_0&=\sum_{t=0}^{T-1}\left[ \mathbf{\hat x}_{t+1}^\top\boldsymbol\Sigma_e^{-1}q(\mathbf{\hat x}_t)+\int_\Omega\boldsymbol \phi^\top(\mathbf r') \mathbf M_{t+1} \boldsymbol\Sigma_e^{-1}  \boldsymbol{\Psi}(\mathbf{r}') f'(\boldsymbol \phi^\top(\mathbf r')\mathbf {\hat x}_t) d\mathbf{r}'\right] \\	
\Xi_1&=\mathbf E_{(\theta',\xi')}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t+1}^\top\right]=\sum_{t=0}^{T-1}\left(\mathbf M_{t+1}+\mathbf{\hat x}_t\mathbf{\hat x}_{t+1}^\top\right) \\

@@

Given the state estimates, $\mathbf {\hat x^{b}}_t$ the state covariance $\mathbf P_t^b$ and cross-covariance $\mathbf M_t^b$ , the $\Xi$-variables can be computed directly.

When taking the derivative of equation~\ref{eq:Voldermont}, the first term described in equation~\ref{eq:def of beta} will disappear. Differntiating with respect to $\boldsymbol\theta$ we have
\begin{equation}\label{eq:Qpartialtheta}
 \frac{\partial \mathcal Q}{\partial \boldsymbol\theta}=2\Xi_0-\boldsymbol\theta^\top(\Xi_2^\top+\Xi_2)-2\xi\Xi_3
\end{equation}
nothing that $\Xi_2$ is symmetric and equating \ref{eq:Qpartialtheta} to the zero vector we get
\begin{equation}\label{eq:theta}
 \boldsymbol \theta=\Xi_2^{-1}\left(\Xi_0^\top-\xi\Xi_3^\top \right)
\end{equation}
similarly differentiating with respect to $\xi$ we have
\begin{equation}\label{eq:Qpartialxi}
 \frac{\partial \mathcal Q}{\partial \xi}=2tr\left\lbrace \Xi_1 \boldsymbol\Sigma_e^{-1}\right\rbrace -2 \Xi_3 \boldsymbol\theta - 2\xi tr \left\lbrace\Xi_4\boldsymbol\Sigma_e^{-1} \right\rbrace
\end{equation}
substituting \ref{eq:theta} in \ref{eq:Qpartialxi} and equating to zero we get
\begin{equation}
 \xi=\frac{tr\left\lbrace \Xi_1 \boldsymbol\Sigma_e^{-1}\right\rbrace-\Xi_3\Xi_2^{-1}\Xi_0^\top}{tr\left\lbrace \Xi_4 \boldsymbol\Sigma_e^{-1}\right\rbrace-\Xi_3\Xi_2^{-1}\Xi_3^\top}
\end{equation}
\subsection{Discussion}
The second derivative of the $\mathcal Q$-function with respect to $\xi$ is given by
\begin{equation}
 \frac{\partial^2 \mathcal Q}{\partial \xi^2}=-2 tr \left\lbrace\Xi_1\boldsymbol\Sigma_e^{-1} \right\rbrace
\end{equation}
$\Xi_1 $ is positive definite if the state vector is persistently excited and the covariance matrix $\boldsymbol\Sigma_e^{-1}$ is also positive definite therefore the trace of $\Xi_1\boldsymbol\Sigma_w^{-1}$ is guaranteed to be positive so the second derivative is negative and represents a maximum of $\mathcal Q$-function.\\
The second derivative of the $\mathcal Q$-function with respect to $\boldsymbol \theta$ is given by
\begin{equation}
 \frac{\partial^2 \mathcal Q}{\partial \boldsymbol\theta^2}=-2 \Xi_2 
\end{equation}
$\Xi_2$ is positive definite and invertible and the second derivative is negative definite and represents a maximum of $\mathcal Q$- function.
